A pipeline to analyse time-course gene expression data

### 2020-07-15 scripts/manuscript.Rmd
updates

genes_to_keep = row.names(
    log_fold_change_max[
	(rowSums(log_fold_change_max > 2) > 0) &
	(fishers_qval < 0.05), ])
# Keep the data corresponding to the genes of interest in another variable.
y = as.matrix(data[genes_to_keep, ])
```

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(
    moanin_model[genes_to_keep, ], n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

The `scores_and_labels` list contains two elements:

    - `scores`: the matrix of shape `n_cluster` x `n_genes`,
      containing for each gene and each cluster, the goodness of fit as
      described above.
    - `labels`: the labels for all of the genes with a
      sufficiently good goodness-of-fit score.

**Assigning cluster labels:** We could just assign each gene to a cluster based on which cluster gave the minimum score. By default, `splines_kmeans_score_and_label` does not do that, but rather requires a sufficiently low enough goodness-of-fit score. The criteria for being "sufficiently low" is based on looking at the distribution of the scores of all genes on all clusters (i.e. the entire `scores` matrix returned by `splines_kmeans_score_and_label`). A gene is then assigned to a cluster only if their best score is above the 50% percentile of that distribution, with the remaining genes getting `NA` as their assignment. Note that because the distribution of scores of genes across *all* clusters are used, this is not equivalent to assigning 50% of genes to a cluster -- it is possible that all genes are assigned to a cluster.

@@

We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
cluster if we did not have the filtering cutoff of `splines_kmeans_score_and_label`, i.e. if we simply assign every gene to a cluster based on which gives them their minimum score.
We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure.

```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
# This is different from output of `splines_kmeans_score_and_label`, 
# which filters the genes
scores = rowMin(scores_and_labels$scores)
labels = apply(scores_and_labels$scores, 1, which.min)
max_score = quantile(scores, 0.5)
par(mfrow=c(3, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(scores[labels == cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
    abline(v=max_score, col="red", lwd=3, lty=2)
}

@@

**EAP** In fact, `which.min` which you used above, picks the first match, so actually a preference to assign into first cluster in this plot.
**NV** Fair point… maybe this should be fixed somehow by adding a random
perturbation to the scores when the scores are equal? Although I think the
text is still appropriate: those genes should definitely be filtered out.

@@

```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
# order them by their score in cluster
ord = order(scores_and_labels$scores[genes_in_cluster2,
                     cluster_of_interest])
cluster2Score = genes_in_cluster2[ord]
dataToPlot = rbind(data[cluster2Score[1:4], ],
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
plot_splines_data(dataToPlot, moanin_model,
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

data_to_plot = data[genes_to_plot, ]
submeta = meta
ord = order(
  submeta$Group,
  submeta$Timepoint,
  submeta$Replicate)
submeta$Timepoint = as.factor(submeta$Timepoint)

data_to_plot = data_to_plot[, ord]
submeta = submeta[ord, ]

res = aheatmap(
    data_to_plot,
    Colv=NA,
    color="YlGnBu",
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=FALSE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(raw)"),
    treeheight=0, legend=FALSE)

# Now use the results of the previous call to aheatmap to reorder the genes.
aheatmap(
    moanin::rescale_values(data_to_plot)[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)

@@

**EAP:** We shouldn't use any hidden functions (i.e. that use `:::` as in `moanin:::rescale_values`). We should make it a exported function if use it in workflow.

**EAP** actually, why do we need to rescale them between 0-1 rather than just a normal call to `scale`? Would it really make a difference?

@@

i = 1
for(n_cluster in all_possible_n_clusters){
    clustering_results = splines_kmeans(
	y, moanin_model,
	n_clusters=n_cluster, random_seed=42,
	n_init=10)
    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
    all_clustering[[i]] = clustering_results$clusters
    i = i + 1
}
```

@@

random_subsample_y = y[indices, ]
kmeans_clusters = moanin::splines_kmeans(
    random_subsample_y, splines_model, n_clusters=n_clusters,
    random_seed=random_seed,
    n_init=20)

# Perform prediction on the whole set of data.
kmeans_clusters = moanin::: splines_kmeans_predict(
    y, kmeans_clusters)
```

>>>>

genes_to_keep = row.names(
    log_fold_change_max[
	(rowSums(log_fold_change_max > 2) > 0) &
	(fishers_qval < 0.05), ])
# Keep the data corresponding to the genes of interest in another variable.
# by subsetting the `moanin_model`, which contains the data.
y = moanin_model[genes_to_keep,]
```

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(y, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

The `scores_and_labels` list contains three elements:

    - `scores`: the matrix of shape `n_cluster` x `n_genes`,
      containing for each gene and each cluster, the goodness of fit as
      described above.
    - `labels`: the labels for all of the genes with a
      sufficiently good goodness-of-fit score.
    - `score_cutoff`: the cutoff used on `scores` to determine whether to 
      assign a label

**Assigning cluster labels:** We could just assign each gene to a cluster based on which cluster gave the minimum score. By default, `splines_kmeans_score_and_label` does not do that, but rather requires a sufficiently low enough goodness-of-fit score. The criteria for being "sufficiently low" is based on looking at the distribution of the scores of all genes on all clusters (i.e. the entire `scores` matrix returned by `splines_kmeans_score_and_label`). A gene is then assigned to a cluster only if their best score is above the 50% percentile of that distribution, with the remaining genes getting `NA` as their assignment (this choice can be changed by `proportion_genes_to_label`). Note that because the distribution of scores of genes across *all* clusters are used, this is not equivalent to assigning 50% of genes to a cluster -- it is possible that all genes are assigned to a cluster.

@@

We can visualize the impact of this filtering process by considering the distribution of goodness-of-fit scores for each
cluster if we did not have the filtering cutoff of `splines_kmeans_score_and_label`, i.e. if we simply assign every gene to a cluster based on which gives them their minimum score.
We display for each cluster, the scores of the genes assigned to that cluster (Figure \ref{visualizeScores}), as compared to the cutoff value for what that score must be under our filtering procedure. We can do this by rerunning `splines_kmeans_score_and_label` and setting the filter `percentage_genes_to_label=1` (and we can speed up the calculation by giving our previously calculated score via the argument `previous_scores`)

```{r visualizeScores, fig.height=6, fix.width=6, echo=TRUE, fig.cap="Distribution of goodness-of-fit scores for each cluster. The dashed red line indicates the scoring threshold: all genes with a score above this threshold will not be labeled."}
# Get the best score and best label for all of the genes
# This time without filtering labels
# We can give the previous calculated scores to `previous_scores` to save time
unfiltered_scores<-splines_kmeans_score_and_label(moanin_model,kmeans_clusters,proportion_genes_to_label=1, previous_scores=scores_and_labels$scores)
best_score = rowMin(scores_and_labels$scores)
best_label = unfiltered_scores$labels
par(mfrow=c(3, 3))
n_clusters = dim(kmeans_clusters$centroids)[1]
for(cluster_id in 1:n_clusters){
    hist(best_score[best_label==cluster_id],
	 breaks=(1:50/50), xlim=c(0, 1),
	 col="black", main=paste("C", cluster_id, sep=""),
	 xlab="score", ylab="Num. genes")
    abline(v=scores_and_labels$score_cutoff, col="red", lwd=3, lty=2)
}

@@

@@

```{r topScoreInCluster2, fig.height=6, fig.width=6, fig.cap=paste("Top scoring genes in Cluster",cluster_of_interest)}
# order them by their score in cluster
ord = order(scores_and_labels$scores[genes_in_cluster2,
                     cluster_of_interest])
cluster2Score = genes_in_cluster2[ord]
dataToPlot = rbind(data[cluster2Score[1:4], ],
    "Centroid"=kmeans_clusters$centroids[cluster_of_interest, ])
plot_splines_data(moanin_model, data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

# order the observations by Group, then time, then replicate
orderedMeta<-colData(moanin_model)
ord = order(
  submeta$Group,
  submeta$Timepoint,
  submeta$Replicate)
submeta$Timepoint = as.factor(submeta$Timepoint)

submeta = submeta[ord, ]

res = aheatmap(
    assay(moanin_model[genes_to_plot,ord]),
    Colv=NA,
    color="YlGnBu",
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=FALSE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(raw)"),
    treeheight=0, legend=FALSE)

# Now use the results of the previous call to aheatmap to reorder the genes.
aheatmap(
    rescale_values(moanin_model[genes_to_plot,ord])[res$rowInd,],
    Colv=NA,
    Rowv=NA,
    annCol=submeta[,
                c("Group", "Timepoint")],
    annLegend=TRUE,
    annColors=ann_colors,
    main=paste("Cluster", cluster_to_plot, "(rescaled)"),
    treeheight=0)

@@

@@

i = 1
for(n_cluster in all_possible_n_clusters){
    clustering_results = splines_kmeans(y,
	    n_clusters=n_cluster, random_seed=42,
	    n_init=10)
    wss_values[i] = sum(clustering_results$WCSS_per_cluster)
    all_clustering[[i]] = clustering_results$clusters
    i = i + 1
}
```

@@

random_subsample_y = y[indices, ]
kmeans_clusters = splines_kmeans(y[indices,]
    random_subsample_y, splines_model, n_clusters=n_clusters,
    random_seed=random_seed,
    n_init=20)

# Perform prediction on the whole set of data.
kmeans_clusters = splines_kmeans_predict(y, kmeans_clusters, method="distance")
```
